Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-13 manuscript_agu.tex
Update on Overleaf.

\begin{abstract}
Global storm-resolving models (GSRMs) are the upcoming global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). The high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and a station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

@@

Increasing climate model resolution is one way of improving model accuracy of representation of the climate system \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) is progressing to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) in research conditions today and operationally in the forthcoming years. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8~km \cite{satoh2019,stevens2019}. This is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1~August--10~September 2016) and winter (20~January--1~March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

nextGEMS is a project \cite{nextgems} focused on the research and development of GSRMs at multiple modeling centers and universities in Europe. The project also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON; \citeA{hohenegger2023}), the Integrated Forecasting System [IFS; \citeA{ifs2023}], and their ocean components at eddy-resolving resolutions: ICON-O \cite{korn2022} coupled with ICON and Finite-Element/volumE Sea ice-Ocean Model [FESOM; \citeA{wang2014a}] and Nucleus for European modeling of the Ocean [NEMO; \citeA{madec2023}] coupled with IFS. The project has so far produced ICON and IFS simulations with three development versions called Cycle 1--3 and a pre-final version, with a final production version planned by the end of the project. nextGEMS is not the only project developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in development include: Convection-Permitting Simulations With the E3SM Global Atmosphere Model [SCREAM; \citeA{caldwell2021}], Atmospheric Model [NICAM; \citeA{satoh2008}], Unified Model (UM), eXperimental System for High-resolution modeling for Earth-to-Local Domain [X-SHiELD; \citeA{shield}], Action de Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH; \citeA{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed Sphere [FV3, \citeA{lin2004}], the National Aeronautics and Space Administration (NASA) Goddard Earth Observing System global atmospheric model version 5 [GEOS5; \citeA{putman2011}], Model for Prediction Across Scales [MPAS; \citeA{skamarock2012}], and System for Atmospheric Modeling [SAM; \citeA{khairoutdinov2003}].

@@

The Southern Ocean (SO) is known to be a problematic region for climate model biases \cite{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of surface and in situ observations and being a lower priority region for numerical weather prediction (NWP) and climate model development because of its distance from populated areas. Nevertheless, radiation biases and changes over an area of its size have a substantial influence on the global climate \cite{rintoul2011}, such as affecting the Earth radiation balance, ocean heat, and carbon uptake \cite{williams2023}, and the SO is an important part of the global ocean conveyor belt \cite{wang2014b}. In general, marine clouds have a disproportionate effect on top of atmosphere (TOA) SW radiation due to the relatively low albedo of the sea surface. The relative longitudinal symmetry of the SO means that model cloud biases tend to be similar across longitudes.

Here, we conventionally refer to the SO as ocean regions south of 40°S, low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason for this dividing latitude is to split the SO into about two equal zones, as well as the results by \citeA{schuddeboom2021} (Fig.~2b), showing a contrast in CMIP model radiation biases, \citeA{schuddeboom2019} (Fig.~2) and \citeA{kuma2020} (Fig.~3) showing contrasting radiation biases in the Hadley Centre Global Environmental Model, and \citeA{cesana2022} showing contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings of \citeA{niu2024}, however, support a different dividing line of 62°S based on cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5~km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over high-latitude SO and the total cloud fraction underestimation on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48~hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

In general, sea surface temperature (SST) biases in the SO can originate either in the atmosphere \cite{hyder2018}, caused by too much shortwave heating of the surface, too little longwave cooling of the surface, or in the ocean circulation. Interactions of both are also possible, for example, SST affecting clouds and clouds affecting the surface radiation. Using the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference, \citeA{zhang2023} have shown that SST biases have improved in CMIP6 compared to CMIP5, with SST overall increasing in the later CMIP phase. However, over the SO this resulted in an even higher positive bias, especially in the Atlantic Ocean (AO) sector of the SO, increasing by up to 1°C. \citeA{luo2023} identified that the SO SST bias in an ensemble of 18 CMIP6 models originates not from the surface heat and radiation fluxes (using reanalyses as a reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON is developed and maintained jointly by Deutscher Wetterdienst, Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Previous studies have identified substantial large-scale biases in climate model clouds over the SO, affecting sea surface temperature and the Earth’s albedo. Our aim is to quantify how well the GSRM ICON is simulating clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on CALIPSO (neither of which are now operational) are affected by the abovementioned issues, resulting in a strong underestimation of cloud occurrence below 2 km relative to ground-based lidar observations \cite{mcerlich2021}. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and stations as a reference for the model evaluation.

@@

The Vaisala CL51 and CT25K (photos in Fig.~\ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910~nm and 905~nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4~km (CL51), 7.7~km (CL31 emulation mode with 5~m vertical resolution), and 7.5~km (CT25K). The vertical resolution is 10~m (5~m configurable) in CL51 and 30~m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6~s for CL51 on AA15‐16, 16~s for CT25K on MARCUS and MICRE, 36~s for CL51 on RV \emph{Polarstern}, and about 2.37~s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910~nm are affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910~nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section~\ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910~nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.~\ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/example.png}
\caption{
An example of the attenuated volume backscattering coefficient (AVBC) \textbf{(a)} measured by CL51 during 24~hours on the PS81/3 voyage and \textbf{(b)} an equivalent AVBC simulated with the ALCF from ERA5 data during the same time period. The red line identifies the cloud mask determined by the ALCF.
}
\label{fig:example}
\end{figure}

@@

The Lufft CHM~15k (photo in Fig.~\ref{fig:map}c) is a ceilometer operating at a near-infrared wavelength of 1064~nm. The maximum range is 15.4~km; the vertical resolution is 5~m in the near range (up to 150~m) and 15~m above; the sampling (temporal) resolution is 2~s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section~\ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM~15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

The Automatic Lidar and Ceilometer Framework (ALCF) is a ground-based lidar simulator and a tool for processing observed lidar data, supporting various instruments and models \cite{kuma2021}. It performs radiative transfer calculations to derive equivalent lidar AVBC in an atmospheric model, which can then be compared with observed AVBC. For this purpose, it takes the cloud fraction, liquid and ice mass mixing ratio, temperature, and pressure model fields as an input and is run offline (on the model output rather than inside the model code). The lidar simulator in the ALCF is based on the instrument simulator Cloud Feedback Model Intercomparison Project (CFMIP) Observation Simulator Package (COSP) \cite{bodas-salcedo2011}. After AVBC is calculated, a cloud mask, cloud occurrence by height, and the total cloud fraction are determined. The ALCF has been used by several research teams for model and reanalysis evaluation \cite{kuma2020,kremser2021,guyot2022,pei2023,whitehead2023,mcdonald2024a}.

@@

A coupled (atmosphere--ocean) GSRM version of the ICON model is in development at the nextGEMS project \cite{hohenegger2023}. ICON is an exceptionally versatile model, allowing for simulations ranging from coarse-resolution ESM simulations, GSRM simulations, limited area model simulations, to large eddy simulations (LES), for both weather prediction and climate projections. ICON uses the atmospheric component ICON-A \cite{giorgetta2018}, whose physics is derived from ECHAM6 \cite{stevens2013}, and the ocean component ICON-O \cite{korn2022}. Earlier runs of the GSRM ICON from DYAMOND were evaluated by \citeA{mauritsen2022}.

Here, we use a free-running (i.e., the weather situation in the model does not correspond to reality) coupled GSRM simulation made for the purpose of climate projection. nextGEMS has so far produced four cycles of model runs. We used a Cycle 3 run \emph{ngc3028} produced in 2023 \cite{nextgems2023a,nextgems2023b} for a model time period of 20 January 2020 to 22 July 2025, of which we analyzed the period 2021--2024 (inclusive). While a Cycle 4 run was available, we could not use it due to a lack of availability of the necessary variables. The horizontal resolution of ngc3028 is about 5 km. The model output is available on 90 vertical levels and 3-hourly instantaneous temporal resolution.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/ann.pdf}
\caption{
Artificial neural network (ANN) for prediction of precipitation in lidar backscatter. \textbf{(a)} Diagram showing the TensorFlow structure of the ANN, \textbf{(b)} randomly selected example samples of near-surface backscatter in four categories (clear, fog, rain, and snow), as determined by coincident human-performed weather observations, \textbf{(c)} receiver operating characteristic diagram of the ANN, \textbf{(d)} examples of 10-day time series of human-observed (``HUM'') and predicted precipitation based on an ANN trained on all voyages (``ANN'') and all voyages except for the shown voyage (``ANN2'') during three randomly selected voyages with the available data. Here, by ``randomly selected,'' we mean selected from the top of a permutation generated by a pseudo-random number generator to prevent authors' bias in the selection.
}
\label{fig:ann}
\end{figure}

@@

When subsetted by cyclonic and non-cyclonic situations (Fig.~\ref{fig:cloud-occurrence}d, e), we see that the cyclonic situations have a larger amount of observed cloudiness, including peak and total cloud fraction, both by about 7\%. In the cyclonic situations, the model vertical profiles of cloud occurrence compare well with observations, but they peak higher by about 200 m and stronger by about 8\%. The reanalyses still tend to underestimate cloud occurrence above 1 km by about 5\% and near the surface by about 14\%. Non-cyclonic situations are similar to the general case, partially also because they form the majority of cases.

>>>>

\begin{abstract}
Global storm-resolving models (GSRMs) are the upcoming global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and Macquarie Island station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

@@

Increasing climate model resolution is one way of improving the accuracy of the representation of the climate system in models \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) has progressed to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) for research today and will become operational in the future. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8~km \cite{satoh2019,stevens2019}. This resolution is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1~August--10~September 2016) and winter (20~January--1~March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

The nextGEMS project \cite{nextgems} focuses on the research and development of GSRMs at multiple modeling centers and universities in Europe. The project also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON; \citeA{hohenegger2023}), the Integrated Forecasting System [IFS; \citeA{ifs2023}], and their ocean components at eddy-resolving resolutions: ICON-O \cite{korn2022} coupled with ICON and Finite-Element/volumE Sea ice-Ocean Model [FESOM; \citeA{wang2014a}] and Nucleus for European modeling of the Ocean [NEMO; \citeA{madec2023}] coupled with IFS. The project has so far produced ICON and IFS simulations with three development versions called Cycle 1--3 and a pre-final version, with a final production version planned by the end of the project. nextGEMS is not the only project developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in development include: Convection-Permitting Simulations With the E3SM Global Atmosphere Model [SCREAM; \citeA{caldwell2021}], Atmospheric Model [NICAM; \citeA{satoh2008}], Unified Model (UM), eXperimental System for High-resolution modeling for Earth-to-Local Domain [X-SHiELD; \citeA{shield}], Action de Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH; \citeA{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed Sphere [FV3, \citeA{lin2004}], the National Aeronautics and Space Administration (NASA) Goddard Earth Observing System global atmospheric model version 5 [GEOS5; \citeA{putman2011}], Model for Prediction Across Scales [MPAS; \citeA{skamarock2012}], and System for Atmospheric Modeling [SAM; \citeA{khairoutdinov2003}].

@@

The Southern Ocean (SO) is known to be a problematic region for climate model biases \cite{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of surface and in situ observations and being a lower priority region for numerical weather prediction (NWP) and climate model development because of its distance from populated areas. Nevertheless, radiation biases and changes over an area of its size have a substantial influence on the global climate \cite{rintoul2011}, such as affecting the Earth radiation balance, ocean heat, and carbon uptake \cite{williams2023}, and the SO is also an important part of the global ocean conveyor belt \cite{wang2014b}. In general, marine clouds have a disproportionate effect on top of atmosphere (TOA) SW radiation due to the relatively low albedo of the sea surface. The relative longitudinal symmetry of the SO means that model cloud biases tend to be similar across longitudes.

Here, we refer to the SO as ocean regions south of 40°S, low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason for this dividing latitude is to split the SO into about two equal zones, as well as the results by \citeA{schuddeboom2021} (Fig.~2b) which show a contrast in CMIP model radiation biases. \citeA{schuddeboom2019} (Fig.~2) and \citeA{kuma2020} (Fig.~3) also show contrasting radiation biases in the Hadley Centre Global Environmental Model, which is also supported by \citeA{cesana2022} which displays contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings of \citeA{niu2024}, however, support a different dividing line of 62°S based on cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5~km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over the high-latitude SO and an underestimation of the total cloud fraction  on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48~hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

In general, sea surface temperature (SST) biases in the SO can originate either in the atmosphere \cite{hyder2018}, caused by too much shortwave heating of the surface, too little longwave cooling of the surface, or in the ocean circulation. Interactions of both are also possible, for example, SST affecting clouds and clouds affecting the surface radiation. Using the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference, \citeA{zhang2023} have shown that SST biases have improved in CMIP6 compared to CMIP5, with SST overall increasing in CMIP6. However, over the SO this resulted in an even higher positive bias, especially in the Atlantic Ocean (AO) sector of the SO, increasing by up to 1°C. \citeA{luo2023} identified that the SO SST bias in an ensemble of 18 CMIP6 models originates not from the surface heat and radiation fluxes (using reanalyses as a reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON is developed and maintained jointly by Deutscher Wetterdienst, Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Previous studies have identified substantial large-scale biases in climate model clouds over the SO, affecting sea surface temperature and the Earth’s albedo. Our aim is to quantify how well the GSRM ICON simulates clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on CALIPSO (neither of which are now operational) are affected by the above mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km relative to ground-based lidar observations \cite{mcerlich2021}. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and stations as a reference for the model evaluation.

@@

The Vaisala CL51 and CT25K (photos in Fig.~\ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910~nm and 905~nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4~km (CL51), 7.7~km (CL31 emulation mode with 5~m vertical resolution), and 7.5~km (CT25K). The vertical resolution is 10~m (5~m configurable) in CL51 and 30~m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6~s for CL51 on AA15‐16, 16~s for CT25K on MARCUS and MICRE, 36~s for CL51 on RV \emph{Polarstern}, and about 2.37~s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910~nm are both affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910~nm affected more strongly, but we do not expect this to be a significant issue as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the ALCF (Section~\ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to CL51, equivalent calculations as for CL51 were done assuming a wavelength of 910~nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.~\ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/example.png}
\caption{
An example of the attenuated volume backscattering coefficient (AVBC) \textbf{(a)} measured by the CL51 during 24~hours on the PS81/3 voyage and \textbf{(b)} an equivalent AVBC simulated with the ALCF from ERA5 data during the same time period. The red line identifies the cloud mask determined by the ALCF.
}
\label{fig:example}
\end{figure}

@@

The Lufft CHM~15k (photo in Fig.~\ref{fig:map}c) ceilometer operates at a near-infrared wavelength of 1064~nm. The maximum range is 15.4~km; the vertical resolution is 5~m in the near range (up to 150~m) and 15~m above; the sampling (temporal) resolution is 2~s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section~\ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM~15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

The Automatic Lidar and Ceilometer Framework (ALCF) is a ground-based lidar simulator and a tool for processing observed lidar data, supporting various instruments and models \cite{kuma2021}. It performs radiative transfer calculations to derive equivalent lidar AVBC from an atmospheric model, which can then be compared with observed AVBC. For this purpose, it takes the cloud fraction, liquid and ice mass mixing ratio, temperature, and pressure model fields as an input and is run offline (on the model output rather than inside the model code). The lidar simulator in the ALCF is based on the instrument simulator Cloud Feedback Model Intercomparison Project (CFMIP) Observation Simulator Package (COSP) \cite{bodas-salcedo2011}. After AVBC is calculated, a cloud mask, cloud occurrence by height, and the total cloud fraction are determined. The ALCF has been used by several research teams for model and reanalysis evaluation \cite{kuma2020,kremser2021,guyot2022,pei2023,whitehead2023,mcdonald2024a}.

@@

A coupled (atmosphere--ocean) GSRM version of the ICON model is in development as part of the nextGEMS project \cite{hohenegger2023}. ICON is an exceptionally versatile model, allowing for simulations ranging from coarse-resolution ESM simulations, GSRM simulations, limited area model simulations, to large eddy simulations (LES), for both weather prediction and climate projections. ICON uses the atmospheric component ICON-A \cite{giorgetta2018}, whose physics is derived from ECHAM6 \cite{stevens2013}, and the ocean component ICON-O \cite{korn2022}. Earlier runs of the GSRM ICON from DYAMOND were evaluated by \citeA{mauritsen2022}.

Here, we use a free-running (i.e., the weather situation in the model does not correspond to reality) coupled GSRM simulation made for the purpose of climate projection. nextGEMS has so far produced four cycles of model runs. We used a Cycle 3 run \emph{ngc3028} produced in 2023 \cite{nextgems2023a,nextgems2023b} for a model time period of 20 January 2020 to 22 July 2025, of which we analyzed the period 2021--2024 (inclusive). The horizontal resolution of ngc3028 is about 5 km. The model output is available on 90 vertical levels and 3-hourly instantaneous temporal resolution.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/ann.pdf}
\caption{
Artificial neural network (ANN) for prediction of precipitation in lidar backscatter. \textbf{(a)} Diagram showing the TensorFlow structure of the ANN, \textbf{(b)} randomly selected example samples of near-surface backscatter in four categories (clear, fog, rain, and snow), as determined by coincident manual weather observations, \textbf{(c)} receiver operating characteristic diagram of the ANN, \textbf{(d)} examples of 10-day time series of human-observed (``HUM'') and predicted precipitation based on an ANN trained on all voyages (``ANN'') and all voyages except for the shown voyage (``ANN2'') during three randomly selected voyages with the available data. Here, by ``randomly selected,'' we mean selected from the top of a permutation generated by a pseudo-random number generator to prevent authors' bias in the selection.
}
\label{fig:ann}
\end{figure}

@@

When subsetted by cyclonic and non-cyclonic situations (Fig.~\ref{fig:cloud-occurrence}d, e), we see that the cyclonic situations have a larger amount of observed cloudiness, including peak and total cloud fraction, both by about 7\%. In the cyclonic situations, the model vertical profiles of cloud occurrence compare well with observations, but they peak higher by about 200 m and larger by about 8\%. The reanalyses still tend to underestimate cloud occurrence above 1 km by about 5\% and near the surface by about 14\%. Non-cyclonic situations are similar to the general case, partially also because they form the majority of cases.
